Dissecting the causal question underlying the association between cancer and ADRD

### 2021-06-03 06_presentations/melodem2021/index.Rmd
presentantion melodem plus updated dags in manuscript finally yooohooo

---
title: "Toward a clearer understanding of the inverse association between cancer and dementia"
subtitle: ""  
author: 
  - "L. Paloma Rojas-Saunero"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    seal: false
    nature:
      highlightStyle: github
      highlightLines: true
      highlightLanguage: ["r", "css"]
      countIncrementalSlides: FALSE
      ratio: "16:9"
editor_options: 
  chunk_output_type: console
---

@@

load(here::here("03_figs", "plots.RData"))
source(here::here("02_R", "05_plots.R"))
hr_results <- rio::import(here::here("02_R", "hr_results.csv"))
rd_results_ever <- rio::import(here::here("02_R", "rd_results_ever.csv"))
rd_results_tv <- rio::import(here::here("02_R", "rd_results_tv.csv"))
table_results_bounds <- rio::import(here::here("02_R", "table_results_bounds.csv")) %>% 
  select(1,2,rr) %>% 
  filter(Proxy != "Time to cancer",
         model != "IPTW + IPCW") %>% 
  pivot_wider(names_from = model, values_from = rr)

@@

background-image: url(./figs/ser.jpg), url(./figs/erasmus.png)
background-position: 5% 95%, 95% 5%
background-size: 25%, 15%
class: center, middle

@@

## Towards a clearer understanding of the inverse association between cancer and dementia

.right[

L. Paloma Rojas-Saunero, MD

**Epidemiology department**

???
Hello, my name is Paloma Rojas-Saunero, I am a PhD candidate at Erasmus MC and today I will present my work called "towards a clearer understanding of the inverse association between cancer and dementia".

---
### Introduction

<br><br>

- Observational studies have found an inverse association between cancer and dementia.

@@

- Several genetic and molecular mechanisms have been proposed.

@@

- Researchers have raised concerns related to measurement error, unmeasured confounding and selection bias.

???

...with the intention of identifying drug targets to prevent dementia.
However, researchers have raised concerns related to measurement error, unmeasured confounding and selection bias.

But to understand the potential sources and magnitude of bias, we first need to define the causal question which implies defining what is the mechanism of interest behind cancer diagnosis as an exposure.

@@

### Introduction

```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
\node [right of = p, gray!30] (dp) {$D_t$};
\node [right of = dp, gray!30] (pstar) {$P^*_t$};
\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
\draw[->, gray!30] (p) -- (dp);
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
\draw[->, gray!30] (d) -- (y);
\draw[->, gray!30] (c) -- (dp);
\draw[->, gray!30] (c) -- (pstar);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\end{tikzpicture}
```

$P_0$ = Pin1 targeted-drug; $Y_{t+1}$ = Dementia diagnosis

???

For example, in this graph we defined a drug target for Pin1. This is an enzyme that is over-expressed in tumor development, but it's depletion is associated to neurodegeneration.

We can imagine that someday in the future, there will be a drug that targets Pin1 to delay or prevent dementia, and the effectiveness of this drug would be assessed in an RCT.

Since this drug is not available yet, we can at least imagine that Pin1 expression could become a biomarker in a closer future. Once this is available we could measure Pin1 expression in stored blood samples in population based cohort studies.

---
### Pin1 expression as biomarker (Problem 1: _Confounding_)

```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -1cm, xshift = 1cm](l) {$L$};
\node [right of = p, gray!30] (dp) {$D_t$};
\node [right of = dp, gray!30] (pstar) {$P^*_t$};
\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
\draw[->, gray!30] (p) -- (dp);
\draw[->] (l) -- (p);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p) to [out=45,in=135, looseness=0.3] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (d);
\draw[->, gray!30] (d) -- (y);
\draw[->, gray!30] (c) -- (dp);
\draw[->, gray!30] (c) -- (pstar);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\end{tikzpicture}
```

$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $L$ = Shared causes of $P_0$ and $Y_{t+1}$

???
So in this graph we changed P0 to pin1 overexpression. Since this study would be performed within an observational study, Pin1 expression and dementia may share risk factors such as behavioural or environmental, represented as L. Thus to identify the causal relationship we will have to adjust for all L.

Now, since Pin1 is not available we defined a proxy for it.

---
### Cancer diagnosis as the proxy of Pin1 (Problem 2: _Time zero_)

```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
\node [right of = p, gray!30] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P^*_t$};
\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
\draw[->, gray!30] (p) -- (dp);
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
\draw[->, gray!30] (d) -- (y);
\draw[->, gray!30] (c) -- (dp);
\draw[->, gray!30] (c) -- (pstar);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\end{tikzpicture}
```

$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $P^*_t$ = Cancer diagnosis

???
Given that Pin1-expression is present in tumors, and tumors are only measured through diagnosis, cancer diagnosis may act as the proxy for Pin1 over-expression, which is represented as Pstar.
This would mean that if we study the association between pstar and y in the observed data, we assume that the captured effect is only through P.
The challenge here is that there might be a gap in time between our intended exposure and cancer diagnosis. So we need to be careful to adjust for post-baseline covariates of P or mediators between P and Pstar.

---
### Death prior to cancer diagnosis (Problem 3: _Survival bias_)

```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
\node [right of = p] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P^*_t$};
\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
\draw[->] (p) -- (dp);
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->] (dp) -- (pstar);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
\draw[->, gray!30] (d) -- (y);
\draw[->, gray!30] (c) -- (dp);
\draw[->, gray!30] (c) -- (pstar);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\end{tikzpicture}
```

$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $P^*_t$ = Cancer diagnosis; $D_t$ = Death prior to cancer diagnosis

???
One of the mediators between P and Pstar is death prior to cancer diagnosis. Since individuals are at risk of dying from other causes (such as cardiovascular death), we can only measure Pstar in the subset of individuals who have survived long enough to have a cancer diagnosis.

---
### Death prior to cancer diagnosis (Problem 3: _Survival bias_)

```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
\node [right of = p] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P^*_t$};
\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = dp, yshift = 1cm, xshift = 1cm](c) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
\draw[->] (p) -- (dp);
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->] (dp) -- (pstar);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
\draw[->, gray!30] (d) -- (y);
\draw[->] (c) -- (dp);
\draw[->] (c) -- (pstar);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\end{tikzpicture}
```

$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $P^*_t$ = Cancer diagnosis; $D_t$ = Death prior to cancer diagnosis; $C_1$ = Shared causes of $D_t$ and $P^*$

???

And since several risk factors that increase the risk of cancer might also cause death prior to cancer diagnosis, we need to block this backdoor pathways by adjusting for C.

---
### Death prior to dementia diagnosis (Problem 3: _Survival bias_)

```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
\node [right of = p, gray!30, gray!30] (dp) {$D_t$};
\node [right of = dp, gray!30] (pstar) {$P^*_t$};
\node [right of = pstar] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
\draw[->, gray!30] (p) -- (dp);
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->] (p) to [out=+45,in=+135, looseness=0.3] (d);
\draw[->] (d) -- (y);
\draw[->, gray!30] (c) -- (dp);
\draw[->, gray!30] (c) -- (pstar);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\end{tikzpicture}
```

$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $D_{t+1}$ = Death prior to dementia diagnosis

???
Even if we could have measured Pin1 or a drug in everybody at the same time, death is going to play the rol of a competing event, death prior to dementia diagnosis prevents us from observing dementia. This is key because it puts us in a situation where we have to choose an estimand. In this case a total effect that includes all causal pathways, including the ones mediated through dead would not be of interest, since one of the leading causes of death is cancer we would see an inverse association.

---
### Death prior to dementia diagnosis (Problem 3: _Survival bias_)

>>>>

---
title: "Dissecting the causal question underlying the association between cancer and ADRD"
subtitle: "An opportunity to understand sources of bias"  
author: 
  - "L. Paloma Rojas-Saunero"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    seal: false
    nature:
      highlightStyle: github
      highlightLines: true
      highlightLanguage: ["r", "css"]
      countIncrementalSlides: FALSE
      ratio: "16:9"
editor_options: 
  chunk_output_type: console
---

@@

load(here::here("03_figs", "plots.RData"))
source(here::here("02_R", "05_plots.R"))
source(here::here("02_R", "05b_plot_causes_death.R"))
hr_results <- rio::import(here::here("02_R", "hr_results.csv"))
rd_results_ever <- rio::import(here::here("02_R", "rd_results_ever.csv"))
rd_results_tv <- rio::import(here::here("02_R", "rd_results_tv.csv"))
table_results_bounds <- rio::import(here::here("02_R", "table_results_bounds.csv")) %>%
  select(1,2,rr) %>%
  filter(Proxy != "Time to cancer",
         model != "IPTW + IPCW") %>%
  pivot_wider(names_from = model, values_from = rr)

@@

background-image: url(./figs/erasmus.png)
background-position: 5% 95%
background-size: 25%, 15%
class: center, middle

## Dissecting the causal question underlying the association between cancer and dementia

.right[L. Paloma Rojas-Saunero, MD

**Epidemiology department**]

## Introduction

.pull-left[

Observational studies have found a protective effect of cancer diagnosis in the risk of Alzheimer´s disease or related dementias (ADRD).

These studies lead to discussions of repurposing or augmenting current cancer chemotherapeutics for ADRD
]

.pull-right[ 
![](figs/forest_plot.PNG)
]

.footnote[Ospina-Romero M. JAMA Network Open. 2020]

## Cancer diagnosis as an exposure

```{r, fig.align='center', out.width= "40%"}
include_graphics("figs/causal_risks.PNG")

```
.footnote[L. Frain et al. Alzheimer’s & Dementia. 2017]

.pull-left[
![](figs/pin_knockout.PNG)
Driver J. et al. (2015)
]

.pull-right[

![](figs/linna2021.PNG)
<br>
Lanni C. et al. (2021)
]

class: middle, center

## How can identify a causal pathways like Pin1 with real-world data?

## Aims

- To build a causal directed acyclic graph (DAG) which help us connect the causal question to real-world data on cancer and ADRD and spell out the assumptions we need to identify the effect.

- To describe the challenges and how they translate into the analytic decisions.

- To exemplify this process in an application using data from the Rotterdam Study.

---
### Pin1 targeted-drug and ADRD diagnosis

```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P$};
\node [gray!20,above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [gray!20, below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
\node [gray!20, right of = p] (dp) {$D_t$};
\node [gray!20, right of = dp] (p2) {$P_t$};
\node [gray!20, below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
\node [gray!20, right of = p2] (dp2) {$D_{t+1}$};
\node [right of = dp2] (y) {$Y_{t+1}$};
\node [gray!20, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
\draw[->, gray!20] (p) to (pstar);
\draw[->, gray!20] (p2) to (pstar2);
\draw[->, gray!20] (l) -- (p);
\draw[->, gray!20] (l) -- (p2);
\draw[->, gray!20] (l) to (dp);
\draw[->, gray!20] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[->, gray!20] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, gray!20] (pstar) -- (dp);
\draw[->, gray!20] (pstar2) -- (dp2);
\draw[->, gray!20] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[->, gray!20] (dp) -- (p2);
\draw[->, gray!20] (dp2) -- (y);
\draw[->, gray!20] (cd) -- (dp2);
\draw[->, gray!20] (cd) -- (y);
\draw[->, gray!20, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

$P$ = Pin1 targeted-drug; $Y_{t+1}$ = Dementia diagnosis

???
In a setting where this hypothetical Pin1-targeting drug was developed, the best way to understand its effect on dementia risk would be to have a well-conducted randomized trial in which we randomize eligible participants in late midlife (e.g., ages 50-60 years) to receive this drug or not, and closely monitor ADRD over a lengthy follow-up.

---
### Pin1 expression as biomarker

```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_{t-1}$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [gray!20, below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
\node [gray!20, right of = p] (dp) {$D_t$};
\node [gray!20, right of = dp] (p2) {$P_t$};
\node [gray!20, below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
\node [gray!20, right of = p2] (dp2) {$D_{t+1}$};
\node [right of = dp2] (y) {$Y_{t+1}$};
\node [gray!20, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
\draw[->, gray!20] (p) to (pstar);
\draw[->, gray!20] (p2) to (pstar2);
\draw[->] (l) -- (p);
\draw[->, gray!20] (l) -- (p2);
\draw[->, gray!20] (l) to (dp);
\draw[->, gray!20] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, gray!20] (pstar) -- (dp);
\draw[->, gray!20] (pstar2) -- (dp2);
\draw[->, gray!20] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[->, gray!20] (dp) -- (p2);
\draw[->, gray!20] (dp2) -- (y);
\draw[->, gray!20] (cd) -- (dp2);
\draw[->, gray!20] (cd) -- (y);
\draw[->, gray!20, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

$P_{t-1}$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $L$ = Shared causes of $P_{t-1}$ and $Y_{t+1}$

???
Since this drug is not currently available, at best we can use observational data on Pin1 expression measurements. For example, suppose that a biomarker test was available to measure Pin1 and we measured this biomarker from (stored) baseline blood samples in a population based-cohort that recruited participants in late midlife. In the observational setting, confounding could explain the observed association between Pin1 and ADRD.

---
### Cancer diagnosis as the proxy of Pin1

```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_{t-1}$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
\node [gray!20, right of = p] (dp) {$D_t$};
\node [right of = dp] (p2) {$P_t$};
\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
\node [gray!20, right of = p2] (dp2) {$D_{t+1}$};
\node [right of = dp2] (y) {$Y_{t+1}$};
\node [gray!20, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
\draw[->] (p) to (pstar);
\draw[->] (p2) to (pstar2);
\draw[->] (l) -- (p);
\draw[->] (l) -- (p2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, gray!20] (l) to (dp);
\draw[->, gray!20] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[->, gray!20] (dp) -- (p2);
\draw[->, gray!20] (pstar) -- (dp);
\draw[->, gray!20] (pstar2) -- (dp2);
\draw[->, gray!20] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[->, gray!20] (dp2) -- (y);
\draw[->, gray!20] (cd) -- (dp2);
\draw[->, gray!20] (cd) -- (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

$P_{t-1}$ and $P_{t}$= Pin1 over-expression; $P^*_{t-1}$ and $P^*_t$ = Cancer diagnosis; $Y_{t+1}$ = Dementia diagnosis; $L$ = Shared causes;

???
But Pin1 is not an available biomarker for population-based research, so at best we can only rely on a proxy for it. Because Pin1 over-expression is present in tumors, and tumors are only measured through diagnosis, we may use cancer as a proxy. 
However, unlike measuring Pin1 at the same time to all participants cancer diagnosis is a time-varying process. Although this means we would measure the association between cancer diagnosis over time and ADRD in the observed data, we are assuming that the captured effect is only through the pathway that involves Pin1 expression over time.

---
### Time-varying cancer diagnosis and inmortal-time bias

```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_{t-1}$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
\node [right of = p] (dp) {$D_t$};
\node [right of = dp] (p2) {$P_t$};
\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
\node [gray!20, right of = p2] (dp2) {$D_{t+1}$};
\node [right of = dp2] (y) {$Y_{t+1}$};
\node [gray!20, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
\draw[->] (p) to (pstar);
\draw[->] (p2) to (pstar2);
\draw[->] (l) -- (p);
\draw[->] (l) -- (p2);
\draw[->] (l) to (dp);
\draw[->] (dp) -- (p2);
\draw[->, gray!20] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, gray!20] (pstar) -- (dp);
\draw[->, gray!20] (pstar2) -- (dp2);
\draw[->, gray!20] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[->, gray!20] (dp2) -- (y);
\draw[->, gray!20] (cd) -- (dp2);
\draw[->, gray!20] (cd) -- (y);
\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

$P_{t-1}$ and $P_{t}$= Pin1 over-expression; $P^*_{t-1}$ and $P^*_t$ = Cancer diagnosis; $Y_{t+1}$ = Dementia diagnosis; $L$ = Shared causes; $D_t$ = Death prior to $P_{t}$

???
A challenge that arises by defining cancer diagnosis as the proxy for Pin1 is defining the time zero, the time where follow-up and screening for ADRD would begin after having the exposure measured. The eligibility criteria to join the study may not align with the time cancer diagnosis is measured. This situation may introduce inmortal-time bias. A participant needs to survive to have a cancer diagnosis over time.

This problem is avoided by recognizing the time-varying nature of cancer diagnosis. And previous studies have defined cancer as a tv or time dependant exposure, though only adjusted for baseline covariates.

The price we pay with this approximation is that implicitly, this means that Pin1 would over-express at the time of cancer diagnosis, which is biologically implausible.

---
### Death as a competing event of ADRD

--
```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node [gray!20](p) {$P_{t-1}$};
\node [gray!20, above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [gray!20, below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
\node [gray!20, right of = p] (dp) {$D_t$};
\node [right of = dp] (p2) {$P_t$};
\node [gray!20, below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
\node [right of = p2] (dp2) {$D_{t+1}$};
\node [right of = dp2] (y) {$Y_{t+1}$};
\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
\draw[->, gray!20] (p) to (pstar);
\draw[->, gray!20] (l) -- (p);
\draw[->, gray!20] (l) to (dp);
\draw[->, gray!20] (pstar) -- (dp);
\draw[->, gray!20] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[->, gray!20] (dp) -- (p2);
\draw[->, semithick, gray!20] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->, gray!20] (p2) to (pstar2);
\draw[->] (dp2) -- (y);
\draw[->] (cd) -- (dp2);
\draw[->] (cd) -- (y);
\draw[->, gray!20] (l) -- (p2);
\draw[->, gray!20] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[->, gray!20] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, gray!20] (pstar2) -- (dp2);
\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

$P_{t}$= Pin1 targeted drug; $D_{t+1}$ = Death prior to $Y_{t+1}$; $Y_{t+1}$ = Dementia diagnosis; $C$ = shared causes

???
Let's step back for a minute and imagine that we had indeed randomized participants to pin1 drug.
Even in the context of an RCT, death will act as a competing event since it prevents dementia at future time-points. Talk about the total effect.
---
###  Death as a competing event of ADRD

```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node [gray!20](p) {$P_{t-1}$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [gray!20, below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
\node [gray!20, right of = p] (dp) {$D_t$};
\node [right of = dp] (p2) {$P_t$};
\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
\node [right of = p2] (dp2) {$D_{t+1}$};
\node [right of = dp2] (y) {$Y_{t+1}$};
\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
\draw[->, gray!20] (p) to (pstar);
\draw[->, gray!20] (l) -- (p);
\draw[->, gray!20] (l) to (dp);
\draw[->, gray!20] (pstar) -- (dp);
\draw[->, gray!20] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[->, gray!20] (dp) -- (p2);
\draw[->, semithick, gray!20] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->] (p2) to (pstar2);
\draw[->] (dp2) -- (y);
\draw[->] (cd) -- (dp2);
\draw[->] (cd) -- (y);
\draw[->] (l) -- (p2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->] (pstar2) -- (dp2);
\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

???
Talk about the issue with total effect when we look at the cancer dx.
Talk about separable effects and the CDE

---
### Causal structure for the effect of Pin1 in ADRD using cancer diagnosis as a proxy for Pin1

```{tikz, echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_{t-1}$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
\node [right of = p] (dp) {$D_t$};
\node [right of = dp] (p2) {$P_t$};
\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
\node [right of = p2] (dp2) {$D_{t+1}$};
\node [right of = dp2] (y) {$Y_{t+1}$};
\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
\draw[->] (p) to (pstar);
\draw[->] (l) -- (p);
\draw[->] (l) to (dp);
\draw[->] (pstar) -- (dp);
\draw[->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[->] (dp) -- (p2);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->] (p2) to (pstar2);
\draw[->] (dp2) -- (y);
\draw[->] (cd) -- (dp2);
\draw[->] (cd) -- (y);
\draw[->] (l) -- (p2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->] (pstar2) -- (dp2);
\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

---
### And there is more we could add...

<br>
.pull-left[

![](https://media.giphy.com/media/fVeAI9dyD5ssIFyOyM/giphy.gif)
]

@@

- Time-varying confounding loops

- Error terms for cancer diagnosis

- Measurement and error term for ADRD

- Differential or dependent error structures

- Participants between 60 and 70 years old at study entry

- Without history of cancer diagnosis and dementia at study entry

@@

- Cancer diagnosis and dementia diagnosis is measured through follow-up and linkage with national registries.

- Date and cause of death was collected on a weekly basis via municipal population registries

- Data on covariates is measured repeatedly over visits and through linkage to EHR

---
### Methods

We considered 2 scenarios to define Pin1 over-expression:

@@

- "Cancer ever vs. never" to replicate previous studies (such as Driver 2012)

- Pin1 and cancer diagnosis as a time-varying exposure

20 years of follow-up or up to time of death/ADRD

---
### Methods

  + Time-varying cancer diagnosis: weights represent the product of the inverse probability of being diagnosed with cancer over time conditional on time-varying covariate history.

- Inverse probability of censoring weights for death (IPCW)*

  + For an individual still alive in year _t_, weights represent the the product of the inverse probability of surviving in each year prior to _t_, conditional on measured common causes of death and dementia.

.footnote[Rojas-Saunero et al. Choosing questions before methods in dementia research with competing events and causal goals.MedRxiv.2021]

@@

### Methods

- Risk of dementia and risk ratios are estimated with a weighted Kaplan-Meier estimator.
--

- Assessed extreme scenarios of independence and fit upper and lower bounds:

  + Total independence: those who died would never develop dementia (lower bound). Aalen-Johanssen estimator treating death as a competing event.

  + Complete dependency: those who died would have dementia prior to death (upper bound). The KM estimator for a composite outcome
--
